---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
    number_sections: true
---

Author: Simon Hobbs  
Date: 28-Oct-2022

---

```{r}

library(dplyr)
library(ggplot2)
library(stringr)
library(hms)
library(chron)

options(scipen=999)

```


# Loading and preprocessing the data

Read the CSV

```{r}

unzip("activity.zip")
df_data <- read.csv("activity.csv", colClasses = c("integer","character","integer"))
summary(df_data)

```

Create column to represent datetime, and time

```{r}

df_data <- mutate(df_data, datetime = 
           as.POSIXct(
               paste(date, str_pad(interval, 4, pad='0')), 
               format="%Y-%m-%d %H%M", tz="UTC" 
                   )
       )  

df_data$time <- as_hms(df_data$datetime)

```


# What is mean total number of steps taken per day?


```{r}

df_summary <- df_data %>% 
    group_by(date) %>% 
    summarise(daily_steps = sum(steps,na.rm=TRUE))

g <- ggplot(df_summary, aes(daily_steps)) + 
    geom_histogram(bins = 20) +
    labs(title="Histogram of Daily Steps")

print(g)

summary(df_summary)

```

```{r}

mean_steps_per_day <- mean(df_summary$daily_steps)
median_steps_per_day <- median(df_summary$daily_steps)

```

**Mean steps per day = `r mean_steps_per_day`.**

**Median steps per day = `r median_steps_per_day`.**


# What is the average daily activity pattern?

```{r}

df_intervalsummary <- df_data %>% 
    group_by(time) %>% 
    summarise(mean_steps_in_interval = mean(steps, na.rm=TRUE))

summary(df_intervalsummary)

g <- ggplot(df_intervalsummary, aes(as.POSIXct(time), mean_steps_in_interval)) +
    geom_line() +
    scale_x_datetime(name = 'time interval', date_labels = "%H:%M") +
    labs(title="Time series plot of Average Steps taken per Time Interval") +
    labs(y = "Average steps in 5min Interval")

print(g)

```

```{r}

df_temp <- df_intervalsummary[which.max(df_intervalsummary$mean_steps_in_interval),]

interval <- format(df_temp[[1]], "%H:%M")

```

**The 5-minute interval containing the maximum number of steps is `r interval`.  **  

**This interval contains (on average) `r df_temp[[2]]` steps.  ** 


# Imputing missing values

**Q1. The total number of missing entries is `r sum(is.na(df_data$steps))`.** 

Q2. I will impute the missing values, based on the following strategy:  
- For any missing entry of 'steps', set it to the mean value for that 5-min
interval, averaged across the dataset. 

Q3. I will now create the new dataset and call it df_data_imputed:

```{r}

# Create a new dataframe that's a copy of df_data, but it's got an extra column
# 'mean_steps_in_interval', looked up based on the interval (time). 
# We'll use that later. 

df_data_imputed <- merge(df_data, df_intervalsummary, sort=FALSE, all.x=TRUE)

# Find out which rows have NA for steps
missing_row_indices <- which(is.na(df_data_imputed$steps))

# For each row which has NA for steps, use the mean_steps_in_interval value.
for (i in missing_row_indices) {
    df_data_imputed$steps[i] <- df_data_imputed$mean_steps_in_interval[i]
}

# Drop the mean_steps_in_interval column    
df_data_imputed <- df_data_imputed %>% select(!mean_steps_in_interval)

```

Q4. Here's a histogram of the dataset after imputation:

```{r}

df_summary_imputed <- df_data_imputed %>% 
    group_by(date) %>% 
    summarise(daily_steps = sum(steps,na.rm=TRUE))

g <- ggplot(df_summary_imputed, aes(daily_steps)) + 
    geom_histogram(bins = 20) +
    labs(title="Histogram of Daily Steps")

print(g)

summary(df_summary_imputed)

```

Q4 (continued). Now let's calculate & report the mean & median values:


```{r}

mean_steps_per_day_imputed <- mean(df_summary_imputed$daily_steps)
median_steps_per_day_imputed <- median(df_summary_imputed$daily_steps)

```

**Mean steps per day = `r mean_steps_per_day_imputed`.**

**Median steps per day = `r median_steps_per_day_imputed`.**


```{r}

df_table <- data.frame(
    imputation = c("None", "Imputed", "Diff"),
    mean_steps_per_day = c(
        mean_steps_per_day, 
        mean_steps_per_day_imputed, 
        mean_steps_per_day_imputed - mean_steps_per_day),
    median_steps_per_day = c(
        median_steps_per_day, 
        median_steps_per_day_imputed, 
        median_steps_per_day_imputed - median_steps_per_day)
    )


df_table

```

We can see that with Imputation, the mean & median number of steps per day increases.  
This makes sense, since we have added steps in!  

The last row of the table above shows the impact of Imputation on the mean & median.  
Imputation adds `r df_table[[3,2]]` steps to the MEAN, and `r df_table[[3,3]]` to the MEDIAN.


# Are there differences in activity patterns between weekdays and weekends?

I'll now create a factor variable called "daytype" for 'weekday' or 'weekend': 

```{r}

df_data_imputed$daytype <- factor(1*is.weekend(df_data_imputed$datetime), 
                                  labels=c("Weekday","weekend"))

```

I'll now group by daytime, and then average the steps within each 5min interval:


```{r}

df_intervalsummary_imputed <- df_data_imputed %>% 
    group_by(daytype, time) %>% 
    summarise(mean_steps_in_interval = mean(steps, na.rm=TRUE))

summary(df_intervalsummary_imputed)

g <- ggplot(df_intervalsummary_imputed, 
            aes(as.POSIXct(time), mean_steps_in_interval)) +
    facet_grid(daytype ~ .) + 
    geom_line() +
    scale_x_datetime(name = 'time interval', date_labels = "%H:%M") +
    labs(title="Panel Plot comparing Average Steps/Interval on Weekdays vs Weekends")

print(g)

```
